---
title: "Problem Set 1"
author: ""
date: "`r format(Sys.Date())`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```


## Installing R and RStudio
* Install R, a free software environment for statistical computing and graphics from [CRAN](https://cran.r-project.org/), the Comprehensive R Archive Network. It is  recommended to install a precompiled binary distribution for your operating system.
* Install RStudio’s IDE (Integrated Development Environment), a powerful user interface for R. RStudio Desktop is freely [available](https://www.rstudio.com).


## Installing R packages

* A strength of R is that many add on packages are available which extend the capability of R. The "official" packages are hosted at CRAN and can be installed in R/RStudio with command \texttt{install.packages()}. The package \texttt{hdm} which we will use later in the course can be installed by

```{r, eval=FALSE}
install.packages("hdm")
```

* Many packages are collected in so-called [task views](https://cran.r-project.org/web/views/). For Machine Learning a good starting point is [here](https://cran.r-project.org/web/views/MachineLearning.html).

* Often the most current version of packages, but also some packages which are not hosted at CRAN, are hosted at file repositories like R Forge and Github. E.g., the most current version of \texttt{hdm} can be installed from R-Forge by specifying the corresponding repository:

```{r, eval=FALSE}
install.packages("hdm", repos="http://R-Forge.R-project.org")
```

* After installing a package it can be made availabe in the current R seesion with the command
```{r, eval=FALSE}
library(hdm)
```
## Packages for Machine Learning

One of the strength of R is that many useful packages for Machine Learning are available. Some of the most important ones which will also useful during the course of this course are given in the table.

package | description |
--------|-------------------------------------------|
mlr | interface to  a large number of classification and regression techniques|
rpart, tree, party |  tree-structured models for regression, classification and survival analysis|
randomForest | random forests|
nnet | single-hidden-layer neural network|
mboost, gbm | boosting methods|
hdm, lars, glmnet | lasso implementations|

## Finding help in R and on the web

* R has a comprehensive built-in help system. E.g. to get help for the function \texttt{lm} which conducts linear regression, you can use any of the following at the program's command prompt :

```{r, eval=FALSE}
help.start()       # general help
help(lm)           # help about function lm
?lm                # same result
apropos("lm")      # list all functions containing string lm
??lm               # extensive search on all documents containing the string "lm"
example(lm)        # show an example of function lm
RSiteSearch("lm") # search for foo in help manuals and archived mailing lists

```

Moreover, many packages contain introductions called "vignettes".

```{r, eval=FALSE}
# get vignettes on using installed packages
vignette()      # show available vingettes
vignette(package="XYZ") # show vignettes in the package XYZ
```

For information on help search in are can be found on this [stackoverflow question](http://stackoverflow.com/questions/15289995/how-to-get-help-in-r)

* Exercise:
a. Install and load the package "hdm".
b. Which function in this package conducts "lasso" estimation?
c. How is this function used?
d. Find the vignette of the package!

* R is shipped with different manuals where "An Introduction to R" is a good starting point to learn more about R.
Moreover, good sources for help are [stackoverflow](http://stackoverflow.com/questions/tagged/r) and the archive of the R-help list where solutions to many problems can be found.

## Course Material

Material for this course is hosted at 
https://github.com/MartinSpindler/Machine-Learning-in-Econometrics









