---
title: "Problem Set 2 -- Linear Regression and Extensions"
author: ''
date: '`r format(Sys.Date())`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```


## Loading and Exploring the Data Set
For this problem set we will analyze the data set *Boston* which is contained in the library *MASS*. This data set records the median house value (*medv*) for $506$ neighborhoods around Boston. The goal is to predict the variable *medv* using $13$ predictors.

* Load the data set.
* Make yourself familiar with the data.  Hint: *str()*, *names()*, *help()*
* Generate Descriptive statistics. Hint: *summary*, *mean*, *sd*, *var*, *min*, *max*, *median*, *range*,  *quantile*, *fivenum*
* Plot the data, especially the outcome variable *medv* and the variable *lstat*. Hint: *plot*, *hist*, *boxplot*

```{r, include=TRUE}
options(warn=1)
library(MASS)
data(Boston)
?Boston
attach(Boston)
summary(medv)
summary(lstat)
```


## Univariate Linear Regression

* Analyse the relation between *medv* and *lstat* with a linear regression. Hint: *lm()*
* Interpret the results. Hint: *summary*
* Plot the regression line in a graph with the original data points.
* What is the predicted value of *medv* for a region with a *lstat* of $32$?
```{r}
reg1 = lm(medv ~ lstat, data=Boston)
summary(reg1)
#plot(reg1)
plot(medv ~ lstat, data=Boston)
abline(reg1, col="red")
predict(reg1, newdata = list(lstat=32), interval="confidence")
```

## Multivariate Linear Regression
* Fit now a multivariate regression.
* Interpret the results, in particular with a focus on the variable *lstat*.
* Fit a more complex model, e.g. considering interaction effects and higher order polynomials.
```{r}
reg2 = lm(medv ~ ., data=Boston)
summary(reg2)
#plot(reg2)
reg3 = lm(medv ~ poly(lstat,3)+ crim + zn + crim:zn+ (chas + nox + rm)^2, data=Boston)
coef(reg3)
```

## Regression Splines
Now we consider again the relation between *lstat* and *medv*.

* Fit a cubic regression spline to the data! Hint: library *splines* and function *bs()*
* Plot the fitted line!
```{r}
library(splines)
par(mfrow=c(1,1))
fit = lm(medv ~ bs(lstat, knots=c(10,20,30)), data=Boston)
lstat.grid <- seq(from=1.8, to=37.9, by=0.1)
pred = predict(fit, newdata=list(lstat=lstat.grid), se=TRUE)
plot(Boston$lstat, Boston$medv, col="gray")
lines(lstat.grid, pred$fit, lwd=2)
lines(lstat.grid, pred$fit + 2*pred$se, lwd="dashed")
lines(lstat.grid, pred$fit - 2*pred$se, lwd="dashed")
```
* Experiment with different spline specifications! Hint: options *knots* and *df*
```{r, include=F}
fit2 = lm(medv ~ bs(lstat, df=6), data=Boston)
attr(bs(lstat, df=6), "knots")
pred2 = predict(fit2, newdata=list(lstat=lstat.grid), se=TRUE)
plot(Boston$lstat, Boston$medv, col="gray")
lines(lstat.grid, pred2$fit, lwd=2)
lines(lstat.grid, pred2$fit + 2*pred2$se, lwd="dashed")
lines(lstat.grid, pred2$fit - 2*pred2$se, lwd="dashed")
```

```{r}
fit2 = lm(medv ~ bs(lstat, df=20), data=Boston)
attr(bs(lstat, df=20), "knots")
pred2 = predict(fit2, newdata=list(lstat=lstat.grid), se=TRUE)
plot(Boston$lstat, Boston$medv, col="gray")
lines(lstat.grid, pred2$fit, lwd=2)
lines(lstat.grid, pred2$fit + 2*pred2$se, lwd="dashed")
lines(lstat.grid, pred2$fit - 2*pred2$se, lwd="dashed")
```

* Fit a natural spline. Hint: *ns()*
```{r}
fit = lm(medv ~ ns(lstat, knots=c(10,20,30)), data=Boston)
lstat.grid <- seq(from=1.8, to=37.9, by=0.1)
pred = predict(fit, newdata=list(lstat=lstat.grid), se=TRUE)
plot(Boston$lstat, Boston$medv, col="gray")
lines(lstat.grid, pred$fit, lwd=2)
lines(lstat.grid, pred$fit + 2*pred$se, lwd="dashed")
lines(lstat.grid, pred$fit - 2*pred$se, lwd="dashed")
```

* Compare the different specifications!

## Smoothing Splines

* Fit and plot a smoothing spline to the data! Hint: *smooth.spline()*
```{r}
plot(lstat, medv, cex=.5, col="darkgrey")
title("Smoothing Splines")
fit = smooth.spline(lstat, medv, df=16)
fit2 = smooth.spline(lstat, medv, cv=FALSE) # generalized CV
fit2$df
lines(fit, col="red", lwd=2)
lines(fit2, col="blue", lwd=2)
legend("topright", legend=c("16 df", "6.8 df"), col=c("red", "blue"), lty=1, lwd=2, cex=.8)
```









